---
title: | 
  | Script for analysing scRNA-seq of day 21 reprogramming intermediates
author: |
  | John F. Ouyang
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: 
      collapsed: false
fontsize: 12pt
pagetitle: "d21i"
---

# Preamble
### Things to note
- All input files can be downloaded from http://hrpi.ddnetbio.com/
- All input files are assumed to be in the data folder

### Clear workspace and load libraries
```{r setup}
rm(list=ls())
library(data.table)
library(Matrix)
library(reticulate)
library(Seurat)
library(ggplot2)
library(patchwork)
library(RColorBrewer)
library(parallel)
```

### Define colour palettes and gene signatures
```{r define}
colorMedia = c("black","blue","magenta")
names(colorMedia) = c("D21-fm","D21-nr","D21-tr")
colorCluster = c("#FFA54F","#C47F3D","#8B5A2B","#A48C74",
                 "#BEBEBE","#5F5F5F","#000000",
                 "#4169E1","#4682B4","#000080","#A020F0","#DDA0DD","#FF00FF")
names(colorCluster) = c("D21fm1","D21fm2","D21fm3","D21fm4",
                        "D21fm5","D21fm6","D21fm7",
                        "D21nr1","D21nr2","D21nr3","D21tr1","D21tr2","D21tr3")
colorIdentity = c("black","olivedrab3","gold","darkorange",
                  "blue","purple","firebrick3","pink")
names(colorIdentity) = c("fibroblast","mixed","early-primed","primed",
                         "naive","nis","nonReprog1","nonReprog2")
geneSig = fread("data/geneSignatures_suppTable03.tab")
geneSigPet = fread("data/petSignatures_suppTable12.tab")
nCores = 30
```

# IO and Seurat preprocessing
### Read in mtx file (same format as cellranger output)
```{r io}
inpMtx = readMM("data/d21iSC_matrix.mtx.gz")
inpBar = fread("data/d21iSC_barcodes.tsv.gz", header = FALSE)
inpGen = fread("data/d21iSC_features.tsv.gz", header = FALSE)
colnames(inpMtx) = inpBar$V1
rownames(inpMtx) = inpGen$V2
inpMtx = as(inpMtx, "dgCMatrix")
```

### Create Seurat object
```{r create}
seu = CreateSeuratObject(counts = inpMtx)
seu$library = factor(seu$orig.ident, levels = c("D21-fm","D21-nr","D21-tr"))
Idents(seu) = "library"
rm(inpMtx, inpBar, inpGen); gc() # Remove objects to save RAM
```

### Seurat preprocessing
```{r preproc}
seu = NormalizeData(seu, normalization.method = "LogNormalize",
                    scale.factor = 10000, verbose = FALSE)
seu = FindVariableFeatures(seu, selection.method = "vst",
                           nfeatures = 1500, verbose = FALSE)
seu = ScaleData(seu, do.scale = TRUE, verbose = FALSE)
```

# Seurat reduction (PCA/UMAP)
### PCA / elbow plot
```{r pca}
nPCs = 15
seu = RunPCA(seu, npcs = 30, seed.use = 42,
             features = VariableFeatures(object = seu), verbose = FALSE)
ElbowPlot(seu, ndims = 30)
```

### UMAP
```{r umap}
# seu = RunTSNE(seu, dims = 1:nPCs, tsne.method = "FIt-SNE", 
#               perplexity = 30, nthreads = nCores, 
#               fast_tsne_path = "/home/john/bin/fast_tsne", seed.use = 42)
seu = RunUMAP(seu, dims = 1:nPCs, umap.method = "uwot", 
              seed.use = 42, verbose = FALSE)
ggOut = DimPlot(seu, cols = colorMedia)
ggOut + coord_fixed(ratio = diff(range(ggOut$data$UMAP_1)) / 
                      diff(range(ggOut$data$UMAP_2)))
```

# Seurat clustering
### Perform clustering and relabel clusters
```{r clust}
set.seed(42)
seu = FindNeighbors(seu, reduction = "pca", dims = 1:nPCs, verbose = FALSE)
seu = FindClusters(seu, resolution = 0.5, random.seed = 42, verbose = FALSE)
remapCluster = c("D21fm7","D21tr3","D21nr3","D21nr2","D21tr2","D21fm2","D21fm5",
                 "D21fm3","D21fm6","D21tr1","D21fm4","D21nr1","D21fm1")
seu$seurat_clusters = remapCluster[as.numeric(seu$seurat_clusters)]
seu$seurat_clusters = factor(seu$seurat_clusters, levels = names(colorCluster))
ggOut = DimPlot(seu, group.by = "seurat_clusters", cols = colorCluster)
ggOut + coord_fixed(ratio = diff(range(ggOut$data$UMAP_1)) / 
                      diff(range(ggOut$data$UMAP_2)))
```

### Proportion plots
```{r prop}
# ggData = data.frame(table(seu$library, seu$seurat_clusters)) # cell number
ggData = data.frame(prop.table(table(seu$library, seu$seurat_clusters), 
                               margin = 2))
colnames(ggData) = c("library", "cluster", "value")
ggplot(ggData, aes(cluster, value, fill = library)) +
  geom_col() + coord_flip() +
  ylab("Proportion of Cells (%)") + xlab("Clusters") + 
  scale_fill_manual(values = colorMedia) + theme_classic(base_size = 24)
```

# scanpy reduction (diffMap/FDL)
### Load scanpy via reticulate and create adata
```{r scanpy}
sc <- import("scanpy", convert = FALSE)
ad <- import("anndata", convert = FALSE)
scipy <- import(module = 'scipy.sparse', convert = FALSE)
sp_sparse_csc <- scipy$csc_matrix

adMat = seu@assays$RNA@data
adMat = sp_sparse_csc(
  tuple(np_array(adMat@x), np_array(adMat@i), np_array(adMat@p)),
  shape = tuple(adMat@Dim[1], adMat@Dim[2]))
adObs = seu@meta.data
adVar = data.frame(geneName = rownames(seu))
adObsm = list()
adObsm[["X_pca"]] = np_array(seu@reductions$pca@cell.embeddings)
adObsm[["X_umap"]] = np_array(seu@reductions$umap@cell.embeddings)
adata = ad$AnnData(X = adMat$T, obs = adObs, var = adVar, obsm = dict(adObsm))
adata$var_names = rownames(seu)
adata$obs_names = colnames(seu)
```

### Compute diffusion maps
```{r diffMap}
sc$pp$neighbors(adata, n_neighbors = as.integer(30), 
                n_pcs = as.integer(nPCs), random_state = as.integer(42))
sc$tl$diffmap(adata, n_comps = as.integer(20))
oupDR = py_to_r(adata$obsm['X_diffmap'])
rownames(oupDR) = colnames(seu)
colnames(oupDR) = paste0("DC_", 0:(20-1))
oupDR = oupDR[, paste0("DC_", seq(5))]
oupDR[, 2] = - oupDR[, 2] # Flip coordinates
seu[["diffmap"]] = CreateDimReducObject(embeddings = oupDR, key = "DC_", 
                                        assay = DefaultAssay(seu))
ggOut = DimPlot(seu, reduction = "diffmap", group.by = "library", cols = colorMedia)
ggOut + coord_fixed(ratio = diff(range(ggOut$data$DC_1)) / 
                      diff(range(ggOut$data$DC_2)))
```

### Compute FDL
```{r fdl}
sc$tl$draw_graph(adata, layout = "fa", init_pos = "X_umap",
                 n_jobs = as.integer(nCores), random_state = as.integer(42))
oupDR = py_to_r(adata$obsm['X_draw_graph_fa'])
rownames(oupDR) = colnames(seu)
colnames(oupDR) = c("FDL_1","FDL_2")
oupDR = oupDR / 10000
seu[["fdl"]] <- CreateDimReducObject(embeddings = oupDR, key = "FDL_", 
                                     assay = DefaultAssay(seu))
ggOut = DimPlot(seu, reduction = "fdl", group.by = "library", cols = colorMedia)
ggRatio = diff(range(ggOut$data$FDL_1)) / diff(range(ggOut$data$FDL_2))
ggOut + coord_fixed(ratio = ggRatio)
ggOut = DimPlot(seu, reduction = "fdl", group.by = "seurat_clusters", cols = colorCluster)
ggOut + coord_fixed(ratio = ggRatio)
```

### Gene expression on FDL
```{r gene, message=FALSE, fig.height=3}
p1 = FeaturePlot(seu, reduction = "fdl", order = TRUE, features = "GATA3")
p1 = p1 + coord_fixed(ratio = ggRatio) + 
  scale_color_gradientn(colors = c("grey85", brewer.pal(9, "OrRd")))
p2 = FeaturePlot(seu, reduction = "fdl", order = TRUE, features = "NANOG")
p2 = p2 + coord_fixed(ratio = ggRatio) + 
  scale_color_gradientn(colors = c("grey85", brewer.pal(9, "OrRd")))
p1 + p2

p1 = FeaturePlot(seu, reduction = "fdl", order = TRUE, features = "ZIC2")
p1 = p1 + coord_fixed(ratio = ggRatio) + 
  scale_color_gradientn(colors = c("grey85", brewer.pal(9, "OrRd")))
p2 = FeaturePlot(seu, reduction = "fdl", order = TRUE, features = "DNMT3L")
p2 = p2 + coord_fixed(ratio = ggRatio) + 
  scale_color_gradientn(colors = c("grey85", brewer.pal(9, "OrRd")))
p1 + p2
```

# Scoring gene signatures
### Load scoring function
```{r geneS}
# Similar to Seurat's AddModuleScore [Tirosh et al, Science (2016)]
geneScoreSc <- function(geneExpr, genes, nBin = 25, nCtrl = 100, 
                        rngseed = 42, nCores = 4){
  # Bin genes by gene expression
  geneBin = Matrix::rowMeans(geneExpr)
  geneBin = data.table(geneID = names(geneBin), avgExpr = geneBin)
  geneBin$binID = cut(geneBin$avgExpr, nBin, labels = FALSE)
  geneList = genes[genes %in% rownames(geneExpr)]
  oupScores = do.call(
    rbind, mclapply(geneList, function(iGene){
      # Generate control aggregate gene expression
      set.seed(rngseed)
      binUse = geneBin[geneID == iGene]$binID
      nPick = min(nrow(geneBin[binID == binUse]), nCtrl)
      geneCtrl = sample(geneBin[binID == binUse]$geneID, 
                        size = nPick, replace = FALSE)
      # Compute score for each gene
      tmpOut = data.table(sampleID = colnames(geneExpr),
                          geneID = iGene, score = geneExpr[iGene,])
      if(length(geneCtrl) > 1){
        tmpOut$score = tmpOut$score - Matrix::colMeans(geneExpr[geneCtrl,])
      } else {
        tmpOut$score = tmpOut$score - geneExpr[geneCtrl,]
      }
      return(tmpOut)
    }, mc.cores = nCores))
  oupScores = oupScores[,.(aveScore = mean(score)), by = "sampleID"]
  oupSc = oupScores$aveScore
  names(oupSc) = oupScores$sampleID
  oupSc = oupSc[colnames(geneExpr)]
  return(oupSc)
}
```

### Score defined gene signatures and determine identity
```{r geneSig}
for(i in unique(geneSig$module)){
  seu[[i]] = geneScoreSc(seu@assays[["RNA"]]@data, nCores = nCores,
                         geneSig[module == i]$geneName)
}
oupSign = seu[[unique(geneSig$module)]]
seu$identity = colnames(oupSign)[apply(oupSign, 1, which.max)]
seu$identity = factor(seu$identity, levels = unique(geneSig$module))
ggOut = DimPlot(seu, reduction = "fdl", group.by = "identity", cols = colorIdentity)
ggOut + coord_fixed(ratio = ggRatio)
```

### Score TE/EPI/PE signatures
```{r geneSigPet, message=FALSE, fig.height=3}
for(i in c("TE","EPI","PE")){
  seu[[i]] = geneScoreSc(seu@assays[["RNA"]]@data, nCores = nCores,
                         geneSigPet[type == paste0("ALL-",i)]$geneName)
}
p1 = FeaturePlot(seu, reduction = "fdl", features = "TE")
p1 = p1 + coord_fixed(ratio = ggRatio) + 
  scale_color_gradientn(colors = rev(brewer.pal(9, "RdYlBu"))[c(1,1:9,9)])
p2 = FeaturePlot(seu, reduction = "fdl", features = "EPI")
p2 = p2 + coord_fixed(ratio = ggRatio) + 
  scale_color_gradientn(colors = rev(brewer.pal(9, "RdYlBu"))[c(1,1:9,9)])
p1 + p2
```

# Session information
### Save Seurat object
```{r save}
saveRDS(seu, file = "data/seu_d21i.rds")
```

### Python modules
```{r pyInfo}
print(paste0("scanpy: ", sc[["__version__"]]))
print(paste0("anndata: ", ad[["__version__"]]))
```

### R 
```{r sessInfo}
sessionInfo()
```


