---
title: | 
  | Script for analysing bulk RNA-seq of iTSC-related samples
author: |
  | John F. Ouyang
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: 
      collapsed: false
fontsize: 12pt
pagetitle: "tscRNA"
---

# Preamble
### Things to note
- All input files can be downloaded from http://hrpi.ddnetbio.com/
- All input files are assumed to be in the data folder

### Clear workspace and load libraries
```{r setup}
rm(list=ls())
library(data.table)
library(ggplot2)
library(pheatmap)
library(RColorBrewer)
library(edgeR)
```

### Define colour palettes
```{r define}
colorMedia = c("black", "darkorange", "blue", "magenta", "purple", 
               "thistle2", "thistle3", "violetred", "thistle2", 
               "olivedrab2", "olivedrab3", "olivedrab4",
               "tomato2", "tomato3", "tomato4")
names(colorMedia) = c("Fibroblast-D0", "Primed-P10", "t2iLGoY-P10", 
                      "d8iTSC-P10", "d21NiTSC-P10", 
                      "TSC-CT", "TSC-blast", "TSC-BT5", "TSC",
                      "EVT-d8iTSC", "EVT-d21NiTSC", "EVT-TSC",
                      "ST-d8iTSC", "ST-d21NiTSC", "ST-TSC")
```

# IO and library normalization
### Read in coldata and raw counts
```{r io}
tscRNAcolData = fread("data/tscRNA_colData_suppTable14.tab")
tscRNArawCts  = fread("data/tscRNA_rawCts.tab.gz")
tscRNArawMat  = as.matrix(tscRNArawCts[, -c(1:3)])
rownames(tscRNArawMat) = tscRNArawCts$geneName
```

### Calculate log2FPKM
```{r log2FPKM}
dge = DGEList(tscRNArawMat, genes = rownames(tscRNArawMat))
dge$genes$Length = tscRNArawCts$geneLen
dge = calcNormFactors(dge)
tscRNAlogFPKM = rpkm(dge, normalized.lib.sizes = TRUE, log = FALSE)
tscRNAlogFPKM = log2(tscRNAlogFPKM + 1)
```

###  Combine with Okae 2018
```{r okae}
tmp = fread("data/tscExt_OkaeLogFPKM.tab.gz")
tscOkaeLogFPKM = as.matrix(tmp[, -c(1:2)])
rownames(tscOkaeLogFPKM) = tmp$geneName
tmp = intersect(rownames(tscRNAlogFPKM), rownames(tscOkaeLogFPKM))
tmp1 = tscRNAlogFPKM[tmp, tscRNAcolData$sampleID]
tmp2 = tscOkaeLogFPKM[tmp, ]

mergeLogFPKM = cbind(tmp1, tmp2)
tmp2 = data.table(sampleID = colnames(tmp2),
                  group = tstrsplit(colnames(tmp2), "_")[[1]],
                  label = colnames(tmp2))
tmp2$group = gsub("^TSCT", "TSC-CT", tmp2$group)
tmp2$group = gsub("^TSblast", "TSC-blast", tmp2$group)
tmp2$group = gsub("-TSblast|-TSCT", "-TSC", tmp2$group)
tmp2$group = gsub("\\(2D)|\\(3D)", "", tmp2$group)
mergeColData = rbindlist(list(tscRNAcolData, tmp2))
mergeColData$study = c(rep("Liu2020", 23), rep("Okae2018", 21))
```

# Gene barplots / LFC barplot
### Gene barplot
```{r gbarplot}
geneList = c("ANPEP","NANOG","ZIC2","KLF17","TFAP2C","TP63","KRT7",
             "TEAD4", "GATA2", "GATA3", "ELF5")
ggData = mergeColData[group %in% names(colorMedia)[1:7]][, -"study"]
ggData$group = factor(ggData$group, levels = names(colorMedia)[1:7])
for(i in geneList){
  ggData[[i]] = mergeLogFPKM[i, ggData$sampleID]
}
ggData = melt.data.table(ggData, id.vars = c("sampleID","group","label"),
                         variable.name = "gene", value.name = "logFPKM")
ggplot(ggData, aes(group, logFPKM, fill = group)) +
  geom_bar(stat = "summary", fun = "mean") + geom_point() + 
  facet_wrap(~gene, nrow = 1) + scale_fill_manual(values = colorMedia[1:7]) +
  theme_classic(base_size = 12) + theme(axis.text.x = element_blank(), 
                                        legend.position = "bottom")
```

### LFC barplot (ST)
```{r lfcST}
# LFC barplot prep
tmp1 = mergeLogFPKM[, mergeColData[grep("d8iTSC", group)]$sampleID] - 
  rowMeans(mergeLogFPKM[, mergeColData[group == "d8iTSC-P10"]$sampleID])
tmp2 = mergeLogFPKM[, mergeColData[grep("d21NiTSC", group)]$sampleID] - 
  rowMeans(mergeLogFPKM[, mergeColData[group == "d21NiTSC-P10"]$sampleID])
tmp3 = mergeLogFPKM[, mergeColData[study == "Okae2018"]$sampleID] - 
  rowMeans(mergeLogFPKM[, mergeColData[grep("^TSC", group)]$sampleID])
tmpMat = cbind(tmp1, tmp2, tmp3)

# LFC barplot (ST)
geneList = c("CGA","CGB1","CSH1","CSH2","CYP19A1","INHA","PSG1","PSG3")
ggData = mergeColData[group %in% names(colorMedia)[c(4:7,13:15)]]
ggData = ggData[, -"study"]
ggData[grep("^TSC", group)]$group = "TSC"
ggData$group = factor(ggData$group, 
                      levels = c("d8iTSC-P10","ST-d8iTSC","d21NiTSC-P10",
                                 "ST-d21NiTSC","TSC","ST-TSC"))
for(i in geneList){
  ggData[[i]] = tmpMat[i, ggData$sampleID] + 2
}
ggData = melt.data.table(ggData, id.vars = c("sampleID","group","label"),
                         variable.name = "gene", value.name = "logFPKM")
ggplot(ggData, aes(group, logFPKM, fill = group)) +
  geom_bar(stat = "summary", fun = "mean") + geom_point() + 
  facet_wrap(~gene, nrow = 1) + ylab("Log2FC relative to iTSC/TSC") + 
  scale_fill_manual(values = colorMedia[c(4,13,5,14,9,15)]) +
  scale_y_continuous(breaks = c(2,4,6,8,10,12), labels = c(0,2,4,6,8,10)) +
  theme_classic(base_size = 12) + theme(axis.text.x = element_blank(), 
                                        legend.position = "bottom")
```

### LFC barplot (EVT)
```{r lfcEVT}
# LFC barplot (EVT)
geneList = c("ADAM12","DIO2","EBI3","FLT1","ITGA1","HTRA4","MCAM","MMP2")
ggData = mergeColData[group %in% names(colorMedia)[c(4:7,10:12)]]
ggData = ggData[, -"study"]
ggData[grep("^TSC", group)]$group = "TSC"
ggData$group = factor(ggData$group, 
                      levels = c("d8iTSC-P10","EVT-d8iTSC","d21NiTSC-P10",
                                 "EVT-d21NiTSC","TSC","EVT-TSC"))
for(i in geneList){
  ggData[[i]] = tmpMat[i, ggData$sampleID] + 2
}
ggData = melt.data.table(ggData, id.vars = c("sampleID","group","label"),
                         variable.name = "gene", value.name = "logFPKM")
ggplot(ggData, aes(group, logFPKM, fill = group)) +
  geom_bar(stat = "summary", fun = "mean") + geom_point() + 
  facet_wrap(~gene, nrow = 1) + ylab("Log2FC relative to iTSC/TSC") + 
  scale_fill_manual(values = colorMedia[c(4,10,5,11,9,12)]) +
  scale_y_continuous(breaks = c(2,4,6,8,10,12), labels = c(0,2,4,6,8,10)) +
  theme_classic(base_size = 12) + theme(axis.text.x = element_blank(), 
                                        legend.position = "bottom")
```

# Correlation comparison with literature
### IO Dong and merge
```{r dong}
tmp = fread("data/tscExt_DongLogFPKM.tab.gz")
colnames(tmp) = c("geneID","geneName","ESC-primed_AN","ESC-primed_H9",
                  "ESC-naive_AN","ESC-naive_H9","TSC-naive_WIBR3",
                  "TSC-naive_ANr1","TSC-naive_ANr2","TSC-naive_H9r1",
                  "TSC-naive_H9r2","TSC-BT5_rep1","TSC-BT5_rep2",
                  "EVT-nTSC_AN","EVT-nTSC_H9","ST-nTSC_AN","ST-nTSC_H9")
tscDongLogFPKM = as.matrix(tmp[, -c(1:2)])
rownames(tscDongLogFPKM) = tmp$geneName
tmp = intersect(rownames(mergeLogFPKM), rownames(tscDongLogFPKM))
tmp1 = mergeLogFPKM[tmp, ]
tmp2 = tscDongLogFPKM[tmp, ]

mergeLogFPKM = cbind(tmp1, tmp2)
tmp2 = data.table(sampleID = colnames(tmp2),
                  group = tstrsplit(colnames(tmp2), "_")[[1]],
                  label = colnames(tmp2))
tmp2$study = "Dong2020"
mergeColData = rbindlist(list(mergeColData, tmp2))
```

### Aggregate logFPKM by cell type
```{r aggr}
aggrLogFPKM = matrix(0, nrow = nrow(mergeLogFPKM), 
                     ncol = uniqueN(mergeColData$group))
colnames(aggrLogFPKM) = unique(mergeColData$group)
rownames(aggrLogFPKM) = rownames(mergeLogFPKM)
for(i in colnames(aggrLogFPKM)){
  aggrLogFPKM[,i] = rowMeans(
    2 ^ mergeLogFPKM[, mergeColData[group == i]$sampleID] - 1)
}
aggrLogFPKM = log2(aggrLogFPKM + 1)
aggrColData = data.frame(
  study = c(rep("Liu2020", 9), rep("Okae2018", 7), rep("Dong2020", 6)))
rownames(aggrColData) = colnames(aggrLogFPKM)
```

### Integrate between datasets and plot Spearman correlation
```{r spearHeat, fig.height=6}
# Integrate between datasets and take Spearman
aggrCor = removeBatchEffect(aggrLogFPKM, batch = aggrColData$study)
aggrCor = cor(aggrCor, method = "spearman")

# Plot correlation heatmap
tmpCol = c("purple", "thistle2", "violetred")
names(tmpCol) = c("Liu2020", "Okae2018", "Dong2020")
pheatmap(aggrCor, 
         cutree_rows = 5, cutree_cols = 5, clustering_method = "ward.D2",
         annotation_row = aggrColData, annotation_col = aggrColData,
         annotation_colors = list(study = tmpCol))
```



# Session information
### R 
```{r sessInfo}
sessionInfo()
```


